---
title: "About the Shiny Web App for Estimating future Bitcoin profit"
author: "Trenton Potgieter"
date: "Sunday, October 19, 2014"
output: html_document
---
###App Overview

This Shiny App estimates the amount of Bitcoins that wl be mined, based on forcasting what the Bitcoin Network difficuloty will be. To use the App, follow these steps.  

####Using the App

1. Enter the current Hash Rate produced by the mining hardware. (For Example 4.363).  
2. Select wether the number entered is in __kilohashes per second__, __Megahashes per second__, __Gigahases per second__ or __Terahashes per second__.  
3. Click the __Submit__ Button.

After clicking __Submit__, the __Results__ tab displays the forcasted Network Dissiculty for next month, what the projected amount of Bitcoins mined (based on the forecasted Network Difficulty) for the month will be and the total amount of Bitcoins accrued for a two year period.  

####Visualizing the Forecast
To see a plot of the model forecast, click on the __Arima Plot__ tab.

####Forecast Model Output
To see the actual results of the Forecast Model's oputput, click o the __Arima Forecast__ tab.

###App Background

Bitcoin is software that tracks and verifies transactions on a public ledger over a peer-to-peer network. Operations and data associated with Bitcoin are decentralized, meaning they are not performed or stored in one single location. Instead, the Bitcoin network consists of computers across the world that automatically store and relay Bitcoin data to each other. The computers' owners voluntarily choose to use and run the Bitcoin software. Anyone can use and run Bitcoin software. Bitcoins are produced through a process called mining. Mining is the competitive use of computational power to calculate a number that falls within a certain range. The first miner to discover a number "target"[^one] that meets the criteria is rewarded with a set amount of brand new bitcoins (currently 25 bitcoins), the competition then repeats for the discovery of a new number[^two]. In other words, when the number in the block header is equal or lower than the "target", the block is accepted by the network and the process to create a new block begins.

Understanding just how difficult[^three] it is to discover the given number below the given target is the key to determining profitability, or measure of success. More importantly is the ability to estimate future mining profitability by predicting what the network difficulty will be. This paper describes the process of forecasting what the network difficulty will be for a two year period and calculating how the profitability based on the current hashrate of the mining hardware.  

###Methods used in the Shiny App
####Data Collection
The data used for working with the current and previously recorded Network statistics (including Difficulty) is downloaded directly using the R programming language[^four]. from the Bitcoin Network by using the Bitcoin Block Explorer website[^five] which is an open source web tool that allows a user to view information about the blocks, addresses, and transactions created by Bitcoin. 

```{r Initialize, echo = TRUE}
options(scipen = 10)
library(shiny)
library(ggplot2)
library(xts)
library(forecast)
library(lubridate)

## Pull the data straight from the website
dt <- read.table("http://blockexplorer.com/q/nethash", header = FALSE,
                 sep = ",", skip = 15)

```

####Exploratory Data Analysis
Each row within the data is a log containing the following:  
* Time when the block was created (UTC).  
* Decimal target.  
* Difficulty.  
* The average number of hashes it takes to solve a block at this difficulty.  

For the process of analysis, Time and Difficulty values are extracted. Any duplicated values are removed and the data is converted from UTC to a standard time format for easier sub setting into monthly periods.  

```{r Clean, echo = TRUE}
## Extract and convert the Date (in UTC) and Difficulty Value
df <- as.data.frame(as.POSIXct(dt[, 2], origin = '1970-01-01'), format = "%Y/%m/%d")
df[, 2] <- dt[, 5]
names(df) <- c("Date", "Difficulty")
df$Date <- as.Date(as.character(df$Date, format = "%Y-%m-%d"))

## Remove duplicate values 
df <- unique(df)
## Create a new xts data.frame
dfXts <- xts(df$Difficulty, df$Date)

```

####Time-Series Modeling
To model the predicted forecast of the network difficulty, the Arima model is used. Using the auto.arima() function in R uses a variation of the Hyndman and Khandakar[^six] algorithm which combines unit root tests, minimization of the AICc and MLE to obtain an ARIMA model. The algorithm[^seven] follows these steps:  

1. The number of differences d is determined using repeated KPSS tests.
2. The values of __p__ and __q__ are then chosen by minimizing the AICc after differencing the data __d__ times. Rather than considering every possible combination of __p__ and __q__, the algorithm uses a stepwise search to traverse the model space.
    a. The best model (with smallest AICc) is selected from the following four:ARIMA(2,d,2), ARIMA(0,d,0), ARIMA(1,d,0), ARIMA(0,d,1). If __d=0__ then the constant __c__ is included; if __d__ is greater or equal to __1__ then the constant __c__ is set to zero. This is called the "current model".  
    b. Variations on the current model are considered:
      * vary __p__ and/or __q__ from the current model by __+ or - 1__
      * include/exclude __c__ from the current model  
      The best model considered so far (either the current model, or one of these variations) becomes the new current model.  
3. Repeat Step __2(b)__ until no lower AICc can be found.  

```{r Model, echo = TRUE}
## Format to monthly and create the Automatic Forecast model
dfXts <- to.monthly(dfXts)
timeSeries <- as.ts(dfXts[, 4], start = c(2009))
fit <- auto.arima(timeSeries)
fcast <- forecast(fit)

```

###Results

By using the auto.arima() to select the best Arima model, we get a forecast (by month) of the Network Difficulty for the next 24 months, as well as the 80% and 95% prediction intervals for those predictions.  

```{r Results, echop = TRUE}
## Show Model output and plot
fcast
plot(fcast)

```

This data can then be used to estimate future profitability, based on the current hashrate of the mining hardware and the forecasted difficulty by using the following equation[^eight]:  

__BTC earned per day = Block Reward / (Difficulty * 2^32 / Hashrate / seconds in a day)__  

Where:  
1. Block Reward is currently 25.  
2. Difficulty is the current or forecasted difficult.  
3. 2 * 2^32 is a constant.  
4. Hashrate is the current hashrate from the mining equipment.  
5. Seconds in a day is 86400 (1 * 60 * 60 * 24).  

###Conclusion

By using the difficulty forecast, current hashrate of the mining equipment and the profitability equation, the estimated amount of Bitcoins earned per month can be predicted.

```{r results, echo = TRUE}
## Calculating the predicted Hash Rate (assumes forcastng has
## been done "fcast") in a data frame.
## since fcast is it's own object, extract the date in a 
## manipulatable foramt for lubridate.
tmp <- as.data.frame(fcast)
## Unfortunately the dates are as factors
d <- as.data.frame(lapply(row.names(tmp), as.character),
                   stringsAsFactors = FALSE) ##date as character
## Create the final index of dates
iDates <- mdy(d)

## Perform the same as above to extract the index of forecasted hasrates.
iForecast <- tmp[, 1]

## Create the output Table
x <- 4.363 ##sample hashrate TH/s
## Convert to kH/s
x <- x * 1000000000000
Table <- data.frame(Date = as.Date(character()),
                    Difficulty = numeric(),
                    Btc = numeric(),
                    Total = numeric(),
                    stringsAsFactors = FALSE)
names(Table) <- c("Date", "Difficulty Forecast", "Estimated Btc/Month",
                  "Total Accumulated Btc")
## Populate the first row of the Table
Table[1, 1] <- iDates[1]
Table[1, 2] <- iForecast[1]
Table[1, 3] <- 25/(iForecast[1] * 2^32/x/86400) * days_in_month(Table[1,1])[[1]]
Table[1, 4] <- Table[1, 3]

## Apply the above format into a loop
for (i in 2:length(iForecast)) {
        Table[i, 1] <- iDates[i]
        Table[i, 2] <- iForecast[i]
        Table[i, 3] <- 25/(iForecast[i] * 2^32/x/86400) * days_in_month(Table[i, 1])[[1]]
        Table[i, 4] <- Table[i, 3] + Table[i-1, 4]
}

```

Below is an example of the results using __4.363 TH/s__.

```{r sample, echo = TRUE}
Table

```

###References

[^one]: https://en.bitcoin.it/wiki/Target
[^two]: https://bitcoinhelp.net/know/how-bitcoin-works
[^three]: https://en.bitcoin.it/wiki/Difficulty
[^four]: http://www.R-project.org
[^five]: http://blockexplorer.com
[^six]: http://robjhyndman.com/papers/automatic-forecasting/
[^seven]: https://www.otexts.org/fpp/8/7
[^eight]: https://en.bitcoin.it/wiki/Difficulty